---
title: "DataTrans_DatTidy_Lab"
author: "Aaliyah Phillips"
date: "`4/21/2026`"
output: html_document
---

###Data Transformation

##Start by loading library(tidyverse) and library(dplyr) if not done already.

```{r}
library(tidyverse)
library(dplyr)
library(nycflights13)
```


3.2.5 Exercises
In a single pipeline for each condition, find all flights that meet the condition:

1. Had an arrival delay of two or more hours
2. Flew to Houston (IAH or HOU)
3. Were operated by United, American, or Delta
4. Departed in summer (July, August, and September)
5. Arrived more than two hours late but didn’t leave late
6. Were delayed by at least an hour, but made up over 30 minutes in flight
7. Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

8. Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

9. Was there a flight on every day of 2013?

10. Which flights traveled the farthest distance? Which traveled the least distance?

11. Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay)
  )
#1. Had an arrival delay of two or more hours
flights %>%
  filter(arr_delay >= 120)

#2. Flew to Houston (IAH or HOU)
flights %>%
  filter(dest %in% c("IAH", "HOU"))

#3. Were operated by United, American, or Delta
flights %>%
  filter(carrier %in% c("UA", "AA", "DL"))

#4. Departed in summer (July, August, and September)
flights %>%
  filter(month %in% 7:9)

#5. Arrived more than two hours late but didn’t leave late
flights %>%
  filter(arr_delay > 120, dep_delay <= 0)
#6. Were delayed by at least an hour, but made up over 30 minutes in flight
flights %>%
  filter(dep_delay >= 60, dep_delay - arr_delay >= 30)
#7. Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.
flights %>%
  arrange(desc(dep_delay))
flights %>%
  arrange(dep_time)
#8. Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)
flights %>%
  mutate(speed_mph = distance / air_time * 60) %>%
  arrange(desc(speed_mph))
#9. Was there a flight on every day of 2013?  
flights %>%
  count(year, month, day) # Yes
#10. Which flights traveled the farthest distance? Which traveled the least distance?
flights %>%
  arrange(desc(distance)) %>%
  slice(1)
flights %>%
  arrange(distance) %>%
  slice(1)

#11. Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.
print("Yes, because filter() reduces rows first, so arrange() has less work to do.")

```



###Displaying three tables

TB cases in world

```{r}
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <dbl>  <dbl>      <dbl>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```
```{r}
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <dbl> <chr>          <dbl>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ℹ 6 more rows
```
```{r}
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#>   <chr>       <dbl> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```
```{r}
# Compute rate per 10,000
table1 |>
  mutate(rate = cases / population * 10000)


# Compute total cases per year
table1 |> 
  group_by(year) |> 
  summarize(total_cases = sum(cases))


# Visualize changes over time
ggplot(table1, aes(x = year, y = cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country, shape = country)) +
  scale_x_continuous(breaks = c(1999, 2000)) # x-axis breaks at 1999 and 2000
```
###5.2.1 Exercises
For each of the sample tables, describe what each observation and each column represents.

Sketch out the process you’d use to calculate the rate for table2 and table3. You will need to perform four operations:

1.  Extract the number of TB cases per country per year.
2.  Extract the matching population per country per year.
3.  Divide cases by population, and multiply by 10000.
4.  Store back in the appropriate place.

You haven’t yet learned all the functions you’d need to actually perform these operations, but you should still be able to think through the transformations you’d need.


```{r}
# Table 1 Observations: country,  columns: year/country/cases/population
# Table 2 observations: country/year/type, columns:  country/year/count/type
# Table 3 observations: country/year, columns: country/year/rate 

#Table 2
table2_tidy <- table2 %>%
  pivot_wider(
    names_from = type,
    values_from = count
  ) %>%
  mutate(rate = cases / population * 10000)

table2_tidy

```


```{r}
#Table 3
table3_tidy <- table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE) %>%
  mutate(rate = cases / population * 10000)

table3_tidy
```

